---
title: "Truncated Voting"
date: '2023-04-13'
output: 
  html_document:
    code_folding: "hide"
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries, include = FALSE}
library(tidyverse)
library(DT)
library(knitr)
library(gridExtra)
library(cowplot)
```

## Truncated Voting Project

In this project we posed and answered the following questions using our election data:


1. For a fixed truncation level $TL$, what percentage of elections satisfying $TL < n-1$ have the property that the RCV winner when using $TL$ is different from the RCV winner when preferences are not truncated?
2. For $1 \leq k \leq n-1$, what percentage of elections have $k$ different winners as we increase $TL$ from $1$ to $n-1$?
3. As we increase the truncation level, does the likelihood of electing the Condorcet winner increase?

### Preliminaries

We analyzed 1183 elections.

* 105 were American elections
* 1078 were Scottish government elections
  + we removed 12 that had majority candidates

```{r}
files <- list.files("ElectionFiles/", pattern = ".(txt|csv|blt)", recursive = TRUE)
election <- "ElectionFiles/"
files <- str_c(election, files)
AmericanFiles <- files[c(1,5)]
ScottishFiles <- files[2:4]
```

All of the American election files have the following format:

* The ballot column is the voter's ballot which gives their preferences. 
* The Count column is the number of voters who voted with the given preferences.
* The number of candidates in the election is given in the first row of the Num Cands column.
* The candidates are listed in the first row of the Cand List column.  All of these files have candidates listed simply as A, B, C, D, etc.

```{r AmericanElectionSample}
files[[1]]
American1 <- read_csv(files[[1]])
head(American1)
tail(American1)
```

All of the Scottish election files have the following format:

* The first line gives the number of candidates and the number of seats being filled in the election
* The next lines until the last few lines give the Count and the voter preferences in order where 1 is the first candidate, 2 is the second candidate, etc.  Thus the line 9 5 4 3 0 means 9 voters ranked candidate 5 first, candidate 4 second, and candidate 3 third.  Each of the lines end in a 0.
* The last few lines list the names of the candidates and the election.

```{r ScottishElectionSample}
files[[2]]
Scottish1 <- read_csv(files[[2]], col_names = FALSE)
head(Scottish1)
tail(Scottish1, n=12)
```

We used the `Fix_data_function()` to get the Scottish election files in a format we could use.

```{r}
Fix_data_function <- function(file) {
  my_data <- read_lines(file) %>%
    str_subset(".+") # get rid of any empty lines
  #head(my_data)
  #Some of the data files have a line at the end with " ", so if it is there, we need to remove it.
  if (my_data[[length(my_data)]] == " ") { 
    my_data <- my_data[- length(my_data)]
  }
    
  ## remove first and last lines
  first_line <- my_data[[1]] %>%
    str_subset(".+") %>%
    str_split("\\s")
  #first_line
  
  ## Read in the file and get the number of lines for the data and number of seats
  Remove_Lines <- strtoi(first_line[[1]][[1]]) + 2
  ###Num_seats <- strtoi(first_line[[1]][[2]])
  Num_Cands <- Remove_Lines - 2
  
  ## remove first and last few lines in the file.
  my_x <- my_data[-1]
  
  for (i in 1:Remove_Lines) {
    my_x <- my_x[-length(my_x)]
  }
  
  #head(my_x)
  
  ## Remove "" 
  my_x_noBlank <- str_subset(my_x, ".+") %>%
    str_trim()
  
  
  ## Make it into a tibble and separate into vote (int) and ballot columns
  my_df <- as_tibble_col(my_x_noBlank) %>%
    mutate(votes = strtoi(str_trim(str_extract(value, "\\d+\\s"))), # get vote column (integer)
           ballot = str_remove(value, "\\d+\\s"), # get ballot column
           ballot = str_sub(ballot, end = -3), # remove 0 and space at the end of ballot
           newballot = str_replace_all(ballot, "26", LETTERS[26]),
           newballot = str_replace_all(newballot, "25", LETTERS[25]),
           newballot = str_replace_all(newballot, "24", LETTERS[24]),
           newballot = str_replace_all(newballot, "23", LETTERS[23]),
           newballot = str_replace_all(newballot, "22", LETTERS[22]),
           newballot = str_replace_all(newballot, "21", LETTERS[21]),
           newballot = str_replace_all(newballot, "20", LETTERS[20]),
           newballot = str_replace_all(newballot, "19", LETTERS[19]),
           newballot = str_replace_all(newballot, "18", LETTERS[18]),
           newballot = str_replace_all(newballot, "17", LETTERS[17]),
           newballot = str_replace_all(newballot, "16", LETTERS[16]),
           newballot = str_replace_all(newballot, "15", LETTERS[15]),
           newballot = str_replace_all(newballot, "14", LETTERS[14]),
           newballot = str_replace_all(newballot, "13", LETTERS[13]),
           newballot = str_replace_all(newballot, "12", LETTERS[12]),
           newballot = str_replace_all(newballot, "11", LETTERS[11]),
           newballot = str_replace_all(newballot, "10", LETTERS[10]),
           newballot = str_replace_all(newballot, "9", LETTERS[9]),
           newballot = str_replace_all(newballot, "8", LETTERS[8]),
           newballot = str_replace_all(newballot, "7", LETTERS[7]),
           newballot = str_replace_all(newballot, "6", LETTERS[6]),
           newballot = str_replace_all(newballot, "5", LETTERS[5]),
           newballot = str_replace_all(newballot, "4", LETTERS[4]),
           newballot = str_replace_all(newballot, "3", LETTERS[3]),
           newballot = str_replace_all(newballot, "2", LETTERS[2]),
           newballot = str_replace_all(newballot, "1", LETTERS[1]),
           newballot = str_remove_all(newballot, " ")) %>%
    select(votes, newballot) %>%
    transmute(ballot = newballot, Count = votes)
  return(list(my_df, Num_Cands))
}

```

Then we started building the table of results so we could eventually answer the questions.

```{r}
candidate_count <- function(file_number) {
  file_tidy <- Fix_data_function(ScottishFiles[[file_number]])
  return(as.integer(file_tidy[[2]]))
}

cand_count <-  map_int(1:3, candidate_count)
scottish_results <- tibble(ScottishFiles, cand_count)
scottish_results
```

We created the `IRV()` function and the `final_function()` which gives us the winner of the election. 

```{r}
IRV <- function(ranking, eliminate) { 
  ranking <- ranking %>%
    mutate(ballot = str_remove(ballot, eliminate)) %>%
    mutate(first_ranked = substr(ballot, 1,1))
  IRV <- ranking %>%
    group_by(first_ranked) %>%
    filter(first_ranked != "") %>%
    summarize(first_rank_count = sum(Count)) %>%
    mutate(total_count = sum(first_rank_count)) %>%
    mutate(majority_percent = first_rank_count / total_count * 100) %>%
    arrange(first_rank_count)
  w <- max(IRV$majority_percent)
  winner <- tail(IRV, n = 1)[[1]][[1]]
  eliminate <- IRV[[1]][[1]]
  return(list(w, winner, eliminate, ranking))
}
```

```{r}
final_function <- function(file_number) {
  clean <-  Fix_data_function(ScottishFiles[[file_number]]) #Fix_data_function(files[[file_number]])
  file_tidy <- clean[[1]]    #read_csv(csv_files[[file_number]], col_select = c(ballot, Count))
  ranking <- file_tidy %>%
  mutate(first_ranked = substr(ballot, 1,1))
  IRV <- ranking %>%
    group_by(first_ranked) %>%
    summarize(first_rank_count = sum(Count)) %>%
    filter(first_ranked != "") %>%
    mutate(total_count = sum(first_rank_count)) %>%
    mutate(majority_percent = first_rank_count / total_count * 100) %>%
    arrange(first_rank_count)
  w <- max(IRV$majority_percent)
  winner <- tail(IRV, n = 1)[[1]][[1]]
  not_first_ranked <- setdiff(LETTERS, IRV$first_ranked)
  for (i in not_first_ranked) {
     ranking <- ranking %>%
      mutate(ballot = str_remove(ballot, i))
  }
  ranking <- ranking %>%
    mutate(first_ranked = substr(ballot, 1,1))
  eliminate <- IRV[[1]][[1]]

if (w <= 50) {
  output <- IRV(ranking, eliminate)
  w <- output[[1]]
  winner <- output[[2]]
  eliminate <- output[[3]]
  ranking <- output[[4]]

  while (w <= 50) { 
    output <- IRV(ranking, eliminate)
    w <- output[[1]]
    winner <- output[[2]]
    eliminate <- output[[3]]
    ranking <- output[[4]]
    }}
  return(winner) 
  }

```

At this point we could add the winner for each election to our data table.

```{r}
winner <- map_chr(1:3, final_function)
voting_dt <- tibble(ScottishFiles, winner) %>%
  left_join(scottish_results) %>%
  select(ScottishFiles, cand_count, winner)
kable(voting_dt)
```

Next we modified that function to get the `trc_function()` so we could get the winner of the truncated ballot.  We would alter the function for each truncation level, so if we wanted TL=3, we would modify the line `mutate(ballot = str_sub(ballot, end = 4))` and have replace the 4 with 3.  We also had to determine the usable elections for the truncation levels, so we would need to specify the Tlist for each TL.  An election is usable at $TL=n$ if the election has at least $n+2$ candidates.



```{r}
head(voting_dt)
Tlist <- voting_dt %>%
  filter(cand_count >= 6) %>%
  select(ScottishFiles)
Tlist <- Tlist[[1]]

trc_function <- function(file_number) {
  clean <- Fix_data_function(Tlist[[file_number]]) #T9list[[file#]]
  file_tidy <- clean[[1]] #read_csv(file, col_select = c(ballot, Count))
  ranking <- file_tidy %>%
    mutate(ballot = str_sub(ballot, end = 4)) %>%
    mutate(first_ranked = substr(ballot, 1,1))
  IRV <- ranking %>%
    group_by(first_ranked) %>%
    summarize(first_rank_count = sum(Count)) %>%
    filter(first_ranked != "") %>%
    mutate(total_count = sum(first_rank_count)) %>%
    mutate(majority_percent = first_rank_count / total_count * 100) %>%
    arrange(first_rank_count)
  w <- max(IRV$majority_percent)
  winner <- tail(IRV, n = 1)[[1]][[1]]
  not_first_ranked <- setdiff(LETTERS, IRV$first_ranked)
  for (i in not_first_ranked) {
     ranking <- ranking %>%
      mutate(ballot = str_remove(ballot, i))
  }
  ranking <- ranking %>%
    mutate(first_ranked = substr(ballot, 1,1))
  eliminate <- IRV[[1]][[1]]

if (w <= 50) {
  output <- IRV(ranking, eliminate)
  w <- output[[1]]
  winner <- output[[2]]
  eliminate <- output[[3]]
  ranking <- output[[4]]

  while (w <= 50) {
    output <- IRV(ranking, eliminate)
    w <- output[[1]]
    winner <- output[[2]]
    eliminate <- output[[3]]
    ranking <- output[[4]]
    }}
    return(winner) 
}
```

```{r}
T4_winners <- map_chr(1:2, trc_function)
T4table1 <- tibble(Tlist, T4_winners) %>%
  mutate(ScottishFiles = Tlist) %>%
  select(-Tlist)
voting_dt_4a <- left_join(voting_dt, T4table1)
kable(voting_dt_4a)
```

We did this for all of the Scottish and American election files and all the truncation levels.

### Results

Here are all of the election files with the winners at each of the truncation levels.

```{r}
AmericanFiles <- read_csv("Results/FinalAmericanFiles.csv")

'%!in%' <- Negate('%in%')

ElectionsWithMajorityCandidates <- c("eilean-siar22/ward_04_preferenceprofile.txt", "sc-borders22/Ward_10___Preference.csv", "orkney12-ballots/orkney12-03.blt", "eilean-siar12-ballots/eilean-siar12-03.blt", "shetland17/shetland17-06.blt", "sc-borders12-ballots/sc-borders12-10.blt", "sc-borders17/sc-borders17-10.blt", "shetland12-ballots/shetland12-06.blt", "e-duns12-ballots/e-duns12-04.blt", "aberdeenshire17-ballots/aberdeenshire17-16.blt", "dumgal12-ballots/dumgal12-01.blt", "w-lothian12-ballots/w-lothian12-09.blt")

FinalScottishVoting <- read_csv("Results/FinalScottishFiles.csv") %>%
  filter(files %!in% ElectionsWithMajorityCandidates) %>%
  filter(files != "scottish by-elections/Edinburgh 2021 by-election Ward 23 Partick East-Kelvindale_preference_profile_report_.csv")

ALLVotingFiles <- bind_rows(AmericanFiles, FinalScottishVoting)
datatable(ALLVotingFiles)

```

#### Question 1:

For a fixed truncation level $TL$, what percentage of elections satisfying $TL < n-1$ have the property that the RCV winner when using $TL$ is different from the RCV winner when preferences are not truncated?

```{r}
Trc_check <- ALLVotingFiles %>%
  mutate(T1_flag = case_when(
    winner == T1_winners ~ 0,
    winner != T1_winners ~ 1
  ),T2_flag = case_when(
    winner == T2_winners ~ 0,
    winner != T2_winners ~ 1
  ),T3_flag = case_when(
    winner == T3_winners ~ 0,
    winner != T3_winners ~ 1
  ),T4_flag = case_when(
    winner == T4_winners ~ 0,
    winner != T4_winners ~ 1
  ),T5_flag = case_when(
    winner == T5_winners ~ 0,
    winner != T5_winners ~ 1
  ),T6_flag = case_when(
    winner == T6_winners ~ 0,
    winner != T6_winners ~ 1
  ),T7_flag = case_when(
    winner == T7_winners ~ 0,
    winner != T7_winners ~ 1
  ),T8_flag = case_when(
    winner == T8_winners ~ 0,
    winner != T8_winners ~ 1
  ),T9_flag = case_when(
    winner == T9_winners ~ 0,
    winner != T9_winners ~ 1
  ),T10_flag = case_when(
    winner == T10_winners ~ 0,
    winner != T10_winners ~ 1  
  ),T11_flag = case_when(
    winner == T11_winners ~ 0,
    winner != T11_winners ~ 1
  ),T12_flag = case_when(
    winner == T12_winners ~ 0,
    winner != T12_winners ~ 1  
    
))


T1p <- (sum(Trc_check$T1_flag)/sum(!is.na(Trc_check$T1_flag)))*100

T2p <- (sum(Trc_check$T2_flag, na.rm = TRUE)/sum(!is.na(Trc_check$T2_flag)))*100

T3p <- (sum(Trc_check$T3_flag, na.rm = TRUE)/sum(!is.na(Trc_check$T3_flag)))*100

T4p <- (sum(Trc_check$T4_flag, na.rm = TRUE)/sum(!is.na(Trc_check$T4_flag)))*100

T5p <- (sum(Trc_check$T5_flag, na.rm = TRUE)/sum(!is.na(Trc_check$T5_flag)))*100

T6p <- (sum(Trc_check$T6_flag, na.rm = TRUE)/sum(!is.na(Trc_check$T6_flag)))*100

T7p <- (sum(Trc_check$T7_flag, na.rm = TRUE)/sum(!is.na(Trc_check$T7_flag)))*100

T8p <- (sum(Trc_check$T8_flag, na.rm = TRUE)/sum(!is.na(Trc_check$T8_flag)))*100

T9p <- (sum(Trc_check$T9_flag, na.rm = TRUE)/sum(!is.na(Trc_check$T9_flag)))*100

T10p <- (sum(Trc_check$T10_flag, na.rm = TRUE)/sum(!is.na(Trc_check$T10_flag)))*100

T11p <- (sum(Trc_check$T11_flag, na.rm = TRUE)/sum(!is.na(Trc_check$T11_flag)))*100

T12p <- (sum(Trc_check$T12_flag, na.rm = TRUE)/sum(!is.na(Trc_check$T12_flag)))*100

percent_list <- c(T1p,T2p,T3p,T4p,T5p,T6p,T7p,T8p,T9p,T10p,T11p,T12p)
levels <- c("T1","T2","T3","T4","T5", "T6", 'T7',"T8", 'T9',"T10","T11","T12")

Q1Answer <- tibble(TL = levels, percent_different = percent_list)
kable(Q1Answer)
```

#### Question 2:

For $1 \leq k \leq n-1$, what percentage of elections have $k$ different winners as we increase $TL$ from $1$ to $n-1$?

```{r}
Elections <- read_csv("Results/AllElectionsCleanData.csv") %>%
  mutate(NumUniqueWin = factor(NumUniqueWin),
         NumCands = factor(NumCands)) %>%
  filter(allFiles %!in% ElectionsWithMajorityCandidates)

MoreDetailUniqueWinTable <- Elections %>%
  group_by(NumCands) %>%
  count(NumUniqueWin) %>%
  pivot_wider(names_from = NumUniqueWin, values_from = n)

colnames(MoreDetailUniqueWinTable) <- (c("Number_of_Candidate","1 winner", "2 winners","3 winners"))

kable(MoreDetailUniqueWinTable)

candidatetable <- Elections %>% count(NumCands)

ggplot(Elections, aes(x = NumCands)) +
  geom_bar(aes(fill = NumUniqueWin)) +
  geom_text(data = candidatetable, aes(x = NumCands, y = n, label = n), nudge_y = 10) +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "Number of Candidates", y = "Number of Elections", 
       title = "Number of Different Winners as TL Varies", fill = "Number of Winners") +
  theme_bw()

```

If there is a change in the winner as the truncation level changes, that change most often occurs at the TL=1 level, so if we remove the TL=1 level we get the following graph.

```{r plot2}
ElectionsWoT1 <- read_csv("Results/AllElectionsCleanWoT1Data.csv") %>%
  mutate(NumUniqueWin = factor(NumUniqueWin),
         NumCands = factor(NumCands)) %>%
  filter(allFiles != "glasgow12/glasgow12-07.blt") %>%
  filter(allFiles != "scottish by-elections/Edinburgh 2021 by-election Ward 23 Partick East-Kelvindale_preference_profile_report_.csv") %>%
  filter(allFiles %!in% ElectionsWithMajorityCandidates)

ggplot(ElectionsWoT1, aes(x = NumCands)) +
  geom_bar(aes(fill = NumUniqueWin)) +
  geom_text(data = candidatetable, aes(x = NumCands, y = n, label = n), nudge_y = 10) +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "Number of Candidates", y = "Number of Elections", 
       title = "Number of Different Winners Disregarding TL=1", fill = "Number of Winners") +
  theme_bw()
```
