---
title: "Ballot Sampling"
date: '2023-04-13'
output: 
  html_document:
    code_folding: "hide"
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries, include = FALSE}
library(tidyverse)
library(DT)
library(knitr)
library(gridExtra)
library(cowplot)
```

We ran simulations to investigate the possible effects of ballot truncation in data like ours.  For this method, we generated 1000 new elections (pseudoprofiles) for each of our elections by

1. randomly sampling $\min \{ 1001, \mbox{number of voters} \}$ ballots from each election with replacement, and
2. calculating the number of different RCV winners in the pseudoprofile as we vary TL from $1$ to $n-1$.

Then from the 1000 pseudoprofiles, we found the maximum number of different RCV winners as we vary the TL.

### Files

The following files are included in the ElectionFiles folder.

```{r ElectionFiles}
files <- list.files("ElectionFiles/", pattern = ".(txt|csv|blt)", recursive = TRUE)
election <- "ElectionFiles/"
files <- str_c(election, files)

kable(files)
```

All of the American election files have the following format:

* The ballot column is the voter's ballot which gives their preferences. 
* The Count column is the number of voters who voted with the given preferences.
* The number of candidates in the election is given in the first row of the Num Cands column.
* The candidates are listed in the first row of the Cand List column.  All of these files have candidates listed simply as A, B, C, D, etc.

```{r}
files[[1]]
American1 <- read_csv(files[[1]])
head(American1)
tail(American1)
```

All of the Scottish election files have the following format:

* The first line gives the number of candidates and the number of seats being filled in the election
* The next lines until the last few lines give the Count and the voter preferences in order where 1 is the first candidate, 2 is the second candidate, etc.  Thus the line 9 5 4 3 0 means 9 voters ranked candidate 5 first, candidate 4 second, and candidate 3 third.  Each of the lines end in a 0.
* The last few lines list the names of the candidates and the election.

```{r}
files[[2]]
Scottish1 <- read_csv(files[[2]], col_names = FALSE)
head(Scottish1)
tail(Scottish1, n=12)
```

### Functions

* Fix_Scottish_data_function: 
  + Input:  file name
  + Output:  Scottish data in the same format as the American data, Candidates

* Fix_American_data_function:
  + Input:  file name
  + Output:  American data, Candidates
  
```{r}
#########################
# Fix_Scottish_data_function
### input:  file name
### output: clean data frame to match American data, candidates
#########################

Fix_Scottish_data_function <- function(file) {
  my_data <- read_lines(file) %>%
    str_subset(".+") # get rid of any empty lines
  #Some of the data files have a line at the end with " ", so if it is there, we need to remove it.
  if (my_data[[length(my_data)]] == " ") { 
    my_data <- my_data[- length(my_data)]
  }
  
  first_line <- my_data[[1]] %>%
    str_subset(".+") %>%
    str_split("\\s")

  ## Read in the file and get the number of lines for the data and number of seats
  Remove_Lines <- strtoi(first_line[[1]][[1]]) + 2
  ###Num_seats <- strtoi(first_line[[1]][[2]])
  Num_Cands <- Remove_Lines - 2
  
  ## remove first and last few lines in the file.
  my_x <- my_data[-1]
  
  for (i in 1:Remove_Lines) {
    my_x <- my_x[-length(my_x)]
  }
  
  ## Remove "" 
  my_x_noBlank <- str_subset(my_x, ".+") %>%
    str_trim()
  
  
  ## Make it into a tibble and separate into vote (int) and ballot columns
  my_df <- as_tibble_col(my_x_noBlank) %>%
    mutate(votes = strtoi(str_trim(str_extract(value, "\\d+\\s"))), # get vote column (integer)
           ballot = str_remove(value, "\\d+\\s"), # get ballot column
           ballot = str_sub(ballot, end = -3), # remove 0 and space at the end of ballot
           newballot = str_replace_all(ballot, "26", LETTERS[26]),
           newballot = str_replace_all(newballot, "25", LETTERS[25]),
           newballot = str_replace_all(newballot, "24", LETTERS[24]),
           newballot = str_replace_all(newballot, "23", LETTERS[23]),
           newballot = str_replace_all(newballot, "22", LETTERS[22]),
           newballot = str_replace_all(newballot, "21", LETTERS[21]),
           newballot = str_replace_all(newballot, "20", LETTERS[20]),
           newballot = str_replace_all(newballot, "19", LETTERS[19]),
           newballot = str_replace_all(newballot, "18", LETTERS[18]),
           newballot = str_replace_all(newballot, "17", LETTERS[17]),
           newballot = str_replace_all(newballot, "16", LETTERS[16]),
           newballot = str_replace_all(newballot, "15", LETTERS[15]),
           newballot = str_replace_all(newballot, "14", LETTERS[14]),
           newballot = str_replace_all(newballot, "13", LETTERS[13]),
           newballot = str_replace_all(newballot, "12", LETTERS[12]),
           newballot = str_replace_all(newballot, "11", LETTERS[11]),
           newballot = str_replace_all(newballot, "10", LETTERS[10]),
           newballot = str_replace_all(newballot, "9", LETTERS[9]),
           newballot = str_replace_all(newballot, "8", LETTERS[8]),
           newballot = str_replace_all(newballot, "7", LETTERS[7]),
           newballot = str_replace_all(newballot, "6", LETTERS[6]),
           newballot = str_replace_all(newballot, "5", LETTERS[5]),
           newballot = str_replace_all(newballot, "4", LETTERS[4]),
           newballot = str_replace_all(newballot, "3", LETTERS[3]),
           newballot = str_replace_all(newballot, "2", LETTERS[2]),
           newballot = str_replace_all(newballot, "1", LETTERS[1]),
           newballot = str_remove_all(newballot, " ")) %>%
    select(votes, newballot) %>%
    transmute(ballot = newballot, Count = votes)
  return(list(my_df, Num_Cands))
}

#########################
# Fix_American_data_function
### input:  file name
### output: clean data frame to match American data, candidates
#########################

Fix_American_data_function <- function(file) {
  my_data <- read_csv(file, col_select = c(ballot, Count, 'Num Cands'))
  Num_Cands <- my_data[[3]][[1]]
  my_df <- my_data %>%
    select(ballot, Count)
  return(list(my_df, Num_Cands))
}
```

#### Example:

Here is the same Scottish election as before but in our fixed format.

```{r} 
files[[2]]
output <- Fix_Scottish_data_function(files[[2]])
head(output[[1]])
tail(output[[1]])
```

And the number of candidates for this election is:
```{r}
output[[2]]
```


* Scottish_Candidate_function
  + Input:  fileNumber
  + Output:  Number of candidates
  
* American_Candidate_function
  + Input:  fileNumber
  + Output:  Number of candidates

```{r}
#########################
# Scottish_Candidate_function
### input:  fileNumber
### output: Number of Candidates
#########################

Scottish_Candidate_function <- function(fileNumber) {
  num_cands <- as.integer(Fix_Scottish_data_function(ScottishFiles[[fileNumber]])[[2]])
  return(num_cands)
}

#########################
# American_Candidate_function
### input:  fileNumber
### output: Number of Candidates
#########################

American_Candidate_function <- function(fileNum) {
  data <- AmericanFiles[[fileNum]]
  my_data <- read_csv(data, col_select = c(ballot, Count, 'Num Cands'))
  Num_Cands <- as.integer(my_data[[3]][[1]])
  return(Num_Cands)
}
```

This function allows us to get all of the candidates for all of the election files.

#### Example:

We will divide our election files into Scottish files and American files.

```{r}
AmericanFiles <- files[c(1,5)]
ScottishFiles <- files[2:4]
```

Then we can create a tibble of the file names and the number of candidates.

```{r}
NumCands <- map_int(1:3, Scottish_Candidate_function)
ScottishElections <- tibble(ScottishFiles, NumCands) %>%
  transmute(File = ScottishFiles, NumCands = NumCands)

NumCandsAmerican <- map_int(1:2, American_Candidate_function)
AmericanElections <- tibble(AmericanFiles, NumCandsAmerican) %>%
  transmute(File = AmericanFiles, NumCands = NumCandsAmerican)
Elections <- bind_rows(ScottishElections, AmericanElections)
Elections

```


* IRV_function2
  + Input:  election data frame with two columns (ballot, Count), NumCands
  + Output:  election winner
  
```{r}
#########################
# IRV_function2
### input:  election data (ballot, Count), NumCands
### output: winner
#########################

IRV_function2 <- function(my_df, num_cands) {
  Num_Voters <- sum(my_df$Count)
  ###############
  # First Round and
  # Get rid of any candidates who didn't get any 1st place votes
  ###############
  R1 <- my_df %>%
    mutate(first = str_sub(ballot, start = 1, end = 1)) %>%
    group_by(first) %>%
    summarise(totalVotes = sum(Count)) %>%
    mutate(majority = totalVotes/sum(totalVotes)) %>%
    arrange(totalVotes)
  
  MajorityVariable <- max(R1$majority)
  
  if (MajorityVariable <= .5) {
    Candidates <- LETTERS[1:num_cands]
    ElimNoVotes <- setdiff(Candidates, R1$first)
    my_df_1 <- my_df
    for (r in ElimNoVotes) {
      my_df_1 <- my_df_1 %>%
        mutate(ballot = str_remove_all(ballot, r))
    }
    while (MajorityVariable <= .5) {
      Rounds <- my_df_1 %>%
        mutate(first = str_sub(ballot, start = 1, end = 1)) %>%
        filter(first != "") %>%
        group_by(first) %>%
        summarise(totalVotes = sum(Count)) %>%
        mutate(majority = totalVotes/sum(totalVotes)) %>%
        arrange(totalVotes)
      
      MajorityVariable <- max(Rounds$majority)
      if (MajorityVariable > .5) {
        WinTable <- Rounds %>% filter(majority > .5)
        Winner <- WinTable[[1]][[1]]
      }
      Elim <- Rounds[[1]][[1]]
      
      # Get rid of losing candidate
      my_df_1 <- my_df_1 %>%
        mutate(ballot = str_remove_all(ballot, Elim))
    }
  } else {
    WinTable <- R1 %>% filter(majority > .5)
    Winner <- WinTable[[1]][[1]]
  }
  return(Winner)
}
```

#### Example:

We can use our IRV function to get the winner of the Scottish file we fixed.

```{r}
files[[2]]
IRV_function2(output[[1]], output[[2]])
```

* IRV_Trunc_function2
  + Input:  election data frame (ballot, Count), Truncation level TL, number of candidates
  + Output:  winner


```{r}
#########################
# IRV_Trunc_function2
### input:  election data (ballot, Count), TL, num_cands
### output: winner
#########################

IRV_Trunc_function2 <- function(TL, data, num_cands) {
  ## truncate ballots
  my_df <- data  %>%
    mutate(ballot = str_sub(ballot, start = 1, end = TL))
  Winner <- IRV_function2(my_df, num_cands)
  return(Winner)
}
```

#### Example:

We can find the winner of our Scottish election with different truncation levels.

Here is TL = 3
```{r}
files[[2]]
IRV_Trunc_function2(3, output[[1]], output[[2]])
```

* sample_function
  + Input:  election data, sample size, number of candidates, truncation levels
  + Output:  data frame (unique_1, unique_2, unique_3, unique_4, unique_Max, uniqueNoT1_1, uniqueNoT1_2, uniqueNoT1_3, uniqueNoT1_4, uniqueNoT1_Max)
  
  This function gives the distribution of the number of unique winners for the 1000 ballot samples (pseudoprofiles).

```{r}
##################
# sample_function
# input: election data, sample_size, num_candsn, tcheck
# output:  tibble(unique_winners_1, unique_winners_2, unique_winners_3, unique_winners_4,
#   unique_winners_max, unique_winners_noT1_1, unique_winners_noT1_2,
#   unique_winners_noT1_3, unique_winners_noT1_4, unique_winners_noT1_max)
#################

sample_function <- function(my_df, sample_size, num_candsn, tcheck) {
  num_unique_winners <- vector("integer", 1000)
  num_unique_winners_noT1 <- vector("integer", 1000)
  for (j in 1:1000) 
  {
    sample_data <- sample(my_df[[1]], sample_size, replace = TRUE, prob = my_df[[2]])
    sample_data_table <- tibble(sample_data) %>%
      transmute(ballot = sample_data,
                Count = 1)
    All_winners <- sapply(tcheck, IRV_Trunc_function2, sample_data_table, num_candsn)
    All_winners_noT1 <- All_winners[-1]
    num_unique_winners[[j]] <- length(unique(All_winners))
    num_unique_winners_noT1[[j]] <- length(unique(All_winners_noT1))
  }
  unique_winners_1 <- length(which(1 == num_unique_winners))
  unique_winners_2 <- length(which(2 == num_unique_winners))
  unique_winners_3 <- length(which(3 == num_unique_winners))
  unique_winners_4 <- length(which(4 == num_unique_winners))
  unique_winners_max <- max(num_unique_winners)
  unique_winners_noT1_1 <- length(which(1 == num_unique_winners_noT1))
  unique_winners_noT1_2 <- length(which(2 == num_unique_winners_noT1))
  unique_winners_noT1_3 <- length(which(3 == num_unique_winners_noT1))
  unique_winners_noT1_4 <- length(which(4 == num_unique_winners_noT1))
  unique_winners_noT1_max <- max(num_unique_winners_noT1)
  outputAll <- tibble(unique_winners_1, unique_winners_2, unique_winners_3, unique_winners_4,
           unique_winners_max, unique_winners_noT1_1, unique_winners_noT1_2,
           unique_winners_noT1_3, unique_winners_noT1_4, unique_winners_noT1_max)
  return(outputAll)
}
```

* Scottish_full_sample_function
  + Input:  fileNumber
  + Output:  data frame (file name, unique_1, unique_2, unique_3, unique_4, unique_Max, uniqueNoT1_1, uniqueNoT1_2, uniqueNoT1_3, uniqueNoT1_4, uniqueNoT1_Max)

* American_full_sample_function
  + Input:  fileNumber
  + Output:  data frame (file name, unique_1, unique_2, unique_3, unique_4, unique_Max, uniqueNoT1_1, uniqueNoT1_2, uniqueNoT1_3, uniqueNoT1_4, uniqueNoT1_Max)
  
```{r}
##################
# Scottish_full_sample_function
# input: num_file
# output:  tibble(filename, unique_winners_1, unique_winners_2, unique_winners_3, unique_winners_4,
#   unique_winners_max, unique_winners_noT1_1, unique_winners_noT1_2,
#   unique_winners_noT1_3, unique_winners_noT1_4, unique_winners_noT1_max))
#################

Scottish_full_sample_function <- function(num_file)
{
  data <- ScottishFiles[[num_file]]
  print(num_file)
  fixed_data <- Fix_Scottish_data_function(data)
  my_df <- fixed_data[[1]]
  num_voters <- sum(my_df$Count)
  sample_size = min(1001, num_voters)
  num_cands <- fixed_data[[2]]
  truncs_to_check = 1:(num_cands - 1)
  samplesAllData <- sample_function(my_df, sample_size, num_cands, truncs_to_check) %>%
    mutate(file = data) %>%
    select(file, everything())
  return(samplesAllData)
}

##################
# American_full_sample_function
# input: num_file
# output:  tibble(filename, unique_winners_1, unique_winners_2, unique_winners_3, unique_winners_4,
#   unique_winners_max, unique_winners_noT1_1, unique_winners_noT1_2,
#   unique_winners_noT1_3, unique_winners_noT1_4, unique_winners_noT1_max))
#################

American_full_sample_function <- function(num_file)
{
  data <- AmericanFiles[[num_file]]
  print(num_file)
  fixed_data <- Fix_American_data_function(data)
  my_df <- fixed_data[[1]]
  num_voters <- sum(my_df$Count)
  sample_size = min(1001, num_voters)
  num_cands <- fixed_data[[2]]
  truncs_to_check = 1:(num_cands - 1)
  samplesAllData <- sample_function(my_df, sample_size, num_cands, truncs_to_check) %>%
    mutate(file = data) %>%
    select(file, everything())
  return(samplesAllData)
}
```

#### Example:

We will not run the code to create the ballot sampling since that takes a significant amount of time, but the code chunk has the code we would run.


```{r eval = FALSE}
#######
# Make table of samples
######

ScottishSamples <- map_dfr(1:3, Scottish_full_sample_function)
AmericanSamples <- map_dfr(1:2, American_full_sample_function)
Samples <- bind_rows(ScottishSamples, AmericanSamples)
ElectionsWSamples <- left_join(Elections, Samples)
ElectionsWSamples
```

## Results

Here are the results from all of the elections.

```{r}
Samples <- read_csv("Results/AllSampleDataNew.csv") %>%
  mutate(NumCands = factor(NumCands),
         unique_winners_max = factor(unique_winners_max)) %>%
  arrange(NumCands)
datatable(Samples)

candidatetable <- Samples %>% count(NumCands)

```

```{r}
MoreDetailSampleUniqueWinTable <- Samples %>%
  group_by(NumCands) %>%
  count(unique_winners_max) %>%
  pivot_wider(names_from = unique_winners_max, values_from = n)

colnames(MoreDetailSampleUniqueWinTable) <- (c("Number_of_Candidate","1 winner", "2 winners","3 winners", "4 winners"))

kable(MoreDetailSampleUniqueWinTable)

```

Below is the graph of our data and the sample data.

```{r plot3}
ggplot(Samples, aes(x = NumCands)) +
  geom_bar(aes(fill = factor(unique_winners_max))) +
  geom_text(data = candidatetable, aes(x = NumCands, y = n, label = n), nudge_y = 10) +
  scale_fill_brewer(palette = "Dark2") +
  scale_x_discrete(limits = candidatetable$NumCands) +
  labs(x = "Number of Candidates", y = "Number of Elections", 
       title = "Ballot Sampling Results", fill = "Number of Unique Winners") +
  theme_bw()

```

Side-by-side plot:

```{r plot4}
plot1 <- ggplot(Samples, aes(x = factor(NumCands))) +
  geom_bar(aes(fill = factor(NumUniqueWin))) +
  geom_text(data = candidatetable, aes(x = factor(NumCands), y = n, label = n), nudge_y = 10) +
  scale_x_discrete(limits = candidatetable$NumCands) +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "Number of Candidates", y = "Number of Elections", 
       title = "Number of Different Winners\nas TL Varies") +
  theme_bw() +
  theme(legend.position = "none")

plot2 <- ggplot(Samples, aes(x = factor(NumCands))) +
  geom_bar(aes(fill = factor(unique_winners_max))) +
  geom_text(data = candidatetable, aes(x = factor(NumCands), y = n, label = n), nudge_y = 10) +
  scale_x_discrete(limits = candidatetable$NumCands) +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "Number of Candidates", y = "Number of Elections", 
       title = "Ballot Sampling Results\nas TL Varies", fill = "Number of \n Unique \n Winners") +
  theme_bw() +
  theme(legend.position = c(.8, .75))

plot_grid(plot1, plot2, ncol = 2)
```

Without T1

```{r plotNoT1}
ElectionsWoT1 <- read_csv("Results/AllElectionsCleanWoT1Data.csv") %>%
  mutate(NumUniqueWin = factor(NumUniqueWin),
         NumCands = factor(NumCands)) %>%
  filter(allFiles != "glasgow12/glasgow12-07.blt") %>%
  filter(allFiles != "scottish by-elections/Edinburgh 2021 by-election Ward 23 Partick East-Kelvindale_preference_profile_report_.csv")

plot1NoT1 <- ggplot(ElectionsWoT1, aes(x = NumCands)) +
  geom_bar(aes(fill = factor(NumUniqueWin))) +
  geom_text(data = candidatetable, aes(x = NumCands, y = n, label = n), nudge_y = 10) +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "Number of Candidates", y = "Number of Elections", 
       title = "Number of Different Winners\nDisregarding TL=1", fill = "Number of Winners") +
  theme_bw() +
  theme(legend.position = "none")

plot2NoT1 <- ggplot(Samples, aes(x = factor(NumCands))) +
  geom_bar(aes(fill = factor(unique_winners_noT1_max))) +
  geom_text(data = candidatetable, aes(x = factor(NumCands), y = n, label = n), nudge_y = 10) +
  scale_x_discrete(limits = candidatetable$NumCands) +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "Number of Candidates", y = "Number of Elections", 
       title = "Ballot Sampling Results\nas TL Varies w/o TL=1", fill = "Number of \n Unique \n Winners") +
  theme_bw() +
  theme(legend.position = c(.8, .75))

plot_grid(plot1NoT1, plot2NoT1, ncol = 2)

```

## Single Seat Elections

We had 132 elections in our data that were single winner elections.  When we restrict our data to just those elections, we get the following plots.

```{r plot5}
SingleSeats <- read_csv("Results/SingleSeatsAll.csv") %>% mutate(allFiles = files)
SingleSeatWinners <- semi_join(Samples, SingleSeats)

SingleSeatWinners %>%
  mutate(SampleNumUniqueWin = unique_winners_max) %>%
  select(allFiles, NumCands, AllWins, NumUniqueWin, SampleNumUniqueWin) %>%  
  datatable()

candidatetable1 <- SingleSeatWinners %>% count(NumCands)

plot1s <- ggplot(SingleSeatWinners, aes(x = factor(NumCands))) +
  geom_bar(aes(fill = factor(NumUniqueWin))) +
  geom_text(data = candidatetable1, aes(x = factor(NumCands), y = n, label = n), nudge_y = 2) +
  scale_x_discrete(limits = candidatetable$NumCands) +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "Number of Candidates", y = "Number of Elections", 
       title = "Single Winner Elections\nand Varying Truncation Level") +
  theme_bw() +
  theme(legend.position = "none")

plot2s <- ggplot(SingleSeatWinners, aes(x = factor(NumCands))) +
  geom_bar(aes(fill = factor(unique_winners_max))) +
  geom_text(data = candidatetable1, aes(x = factor(NumCands), y = n, label = n), nudge_y = 2) +
  scale_x_discrete(limits = candidatetable$NumCands) +
  scale_fill_brewer(palette = "Dark2") +
  labs(x = "Number of Candidates", y = "Number of Elections", 
       title = "Single Winner Elections\nBallot Sampling Results", fill = "Number of \n Unique \n Winners") +
  theme_bw() +
  theme(legend.position = c(.8, .7))

plot_grid(plot1s, plot2s, ncol = 2)

```
