---
title: "Ballot Sampling"
author: "Erin Martin"
date: '2023-04-13'
output: 
  html_document:
    code_folding: "hide"
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries, include = FALSE}
library(tidyverse)
library(DT)
library(knitr)
library(gridExtra)
library(cowplot)
```

## Files

The following files are included in the ElectionFiles folder.

```{r ElectionFiles}
files <- list.files("ElectionFiles/", pattern = ".(txt|csv|blt)", recursive = TRUE)
election <- "ElectionFiles/"
files <- str_c(election, files)

kable(files)
```

All of the American election files have the following format:
* The ballot column is the voter's ballot which gives their preferences 
* The Count column is the number of voters who voted with the given preferences.
* The number of candidates in the election is given in the next column.
* The candidates are listed in the last column.  All of these files have candidates listed simply as A, B, C, D, etc.

```{r}
American1 <- read_csv(files[[1]])
head(American1)
tail(American1)
```

All of the Scottish election files have the following format:
* The first line give the number of candidates and the number of seats being filled in the election
* The next lines until the last few lines give the Count and the voter preferences in order where 1 is the first candidate, 2 is the second candidate, etc.  The list of candidates are given at the end of the file.  Thus the line 9 5 4 3 0 means 9 voters ranked candidate 5 first, candidate 4 second, and candidate 3 third.

```{r}
Scottish1 <- read_csv(files[[2]])
head(Scottish1)
tail(Scottish1, n=12)
```

## Fix Data

We wrote a function so the Scottish data was in the same format as the American data.

```{r}
#########################
# Fix_Scottish_data_function
### input:  file name
### output: clean data frame to match American data, candidates
#########################

Fix_Scottish_data_function <- function(file) {
  my_data <- read_lines(file) %>%
    str_subset(".+") # get rid of any empty lines
  #Some of the data files have a line at the end with " ", so if it is there, we need to remove it.
  if (my_data[[length(my_data)]] == " ") { 
    my_data <- my_data[- length(my_data)]
  }
  
  ## remove first and last lines
  first_line <- my_data[[1]] %>%
    str_subset(".+") %>%
    str_split("\\s")
  #first_line
  
  ## Read in the file and get the number of lines for the data and number of seats
  Remove_Lines <- strtoi(first_line[[1]][[1]]) + 2
  ###Num_seats <- strtoi(first_line[[1]][[2]])
  Num_Cands <- Remove_Lines - 2
  
  ## remove first and last few lines in the file.
  my_x <- my_data[-1]
  
  for (i in 1:Remove_Lines) {
    my_x <- my_x[-length(my_x)]
  }
  
  ## Remove "" 
  my_x_noBlank <- str_subset(my_x, ".+") %>%
    str_trim()
  
  
  ## Make it into a tibble and separate into vote (int) and ballot columns
  my_df <- as_tibble_col(my_x_noBlank) %>%
    mutate(votes = strtoi(str_trim(str_extract(value, "\\d+\\s"))), # get vote column (integer)
           ballot = str_remove(value, "\\d+\\s"), # get ballot column
           ballot = str_sub(ballot, end = -3), # remove 0 and space at the end of ballot
           newballot = str_replace_all(ballot, "26", LETTERS[26]),
           newballot = str_replace_all(newballot, "25", LETTERS[25]),
           newballot = str_replace_all(newballot, "24", LETTERS[24]),
           newballot = str_replace_all(newballot, "23", LETTERS[23]),
           newballot = str_replace_all(newballot, "22", LETTERS[22]),
           newballot = str_replace_all(newballot, "21", LETTERS[21]),
           newballot = str_replace_all(newballot, "20", LETTERS[20]),
           newballot = str_replace_all(newballot, "19", LETTERS[19]),
           newballot = str_replace_all(newballot, "18", LETTERS[18]),
           newballot = str_replace_all(newballot, "17", LETTERS[17]),
           newballot = str_replace_all(newballot, "16", LETTERS[16]),
           newballot = str_replace_all(newballot, "15", LETTERS[15]),
           newballot = str_replace_all(newballot, "14", LETTERS[14]),
           newballot = str_replace_all(newballot, "13", LETTERS[13]),
           newballot = str_replace_all(newballot, "12", LETTERS[12]),
           newballot = str_replace_all(newballot, "11", LETTERS[11]),
           newballot = str_replace_all(newballot, "10", LETTERS[10]),
           newballot = str_replace_all(newballot, "9", LETTERS[9]),
           newballot = str_replace_all(newballot, "8", LETTERS[8]),
           newballot = str_replace_all(newballot, "7", LETTERS[7]),
           newballot = str_replace_all(newballot, "6", LETTERS[6]),
           newballot = str_replace_all(newballot, "5", LETTERS[5]),
           newballot = str_replace_all(newballot, "4", LETTERS[4]),
           newballot = str_replace_all(newballot, "3", LETTERS[3]),
           newballot = str_replace_all(newballot, "2", LETTERS[2]),
           newballot = str_replace_all(newballot, "1", LETTERS[1]),
           newballot = str_remove_all(newballot, " ")) %>%
    select(votes, newballot) %>%
    transmute(ballot = newballot, Count = votes)
  return(list(my_df, Num_Cands))
}
```

Here is the same Scottish election as before but in our fixed format.

```{r} 
output <- Fix_Scottish_data_function(files[[2]])
head(output[[1]])
tail(output[[1]])
output[[2]]
```

Since the formats for the different elections are very different, we will separate the files into American and Scottish.

```{r}
AmericanFiles <- files[c(1,5)]
ScottishFiles <- files[2:4]
```

```{r}
#########################
# Scottish_Candidate_function
### input:  fileNumber
### output: Number of Candidates
#########################

Scottish_Candidate_function <- function(fileNumber) {
  num_cands <- as.integer(Fix_data_function(files[[fileNumber]])[[2]])
  return(num_cands)
}

#########################
# IRV_function2
### input:  election data (ballot, Count), NumCands
### output: winner
#########################

IRV_function2 <- function(my_df, num_cands) {
  Num_Voters <- sum(my_df$Count)
  ###############
  # First Round and
  # Get rid of any candidates who didn't get any 1st place votes
  ###############
  R1 <- my_df %>%
    mutate(first = str_sub(ballot, start = 1, end = 1)) %>%
    group_by(first) %>%
    summarise(totalVotes = sum(Count)) %>%
    mutate(majority = totalVotes/sum(totalVotes)) %>%
    arrange(totalVotes)
  
  MajorityVariable <- max(R1$majority)
  
  if (MajorityVariable <= .5) {
    Candidates <- LETTERS[1:num_cands]
    ElimNoVotes <- setdiff(Candidates, R1$first)
    my_df_1 <- my_df
    for (r in ElimNoVotes) {
      my_df_1 <- my_df_1 %>%
        mutate(ballot = str_remove_all(ballot, r))
    }
    while (MajorityVariable <= .5) {
      Rounds <- my_df_1 %>%
        mutate(first = str_sub(ballot, start = 1, end = 1)) %>%
        filter(first != "") %>%
        group_by(first) %>%
        summarise(totalVotes = sum(Count)) %>%
        mutate(majority = totalVotes/sum(totalVotes)) %>%
        arrange(totalVotes)
      
      MajorityVariable <- max(Rounds$majority)
      if (MajorityVariable > .5) {
        WinTable <- Rounds %>% filter(majority > .5)
        Winner <- WinTable[[1]][[1]]
      }
      Elim <- Rounds[[1]][[1]]
      
      # Get rid of losing candidate
      my_df_1 <- my_df_1 %>%
        mutate(ballot = str_remove_all(ballot, Elim))
    }
  } else {
    WinTable <- R1 %>% filter(majority > .5)
    Winner <- WinTable[[1]][[1]]
  }
  return(Winner)
}

#########################
# IRV_Trunc_function2
### input:  election data (ballot, Count), TL, num_cands
### output: winner
#########################

IRV_Trunc_function2 <- function(TL, data, num_cands) {
  ## truncate ballots
  my_df <- data  %>%
    mutate(ballot = str_sub(ballot, start = 1, end = TL))
  Winner <- IRV_function2(my_df, num_cands)
  return(Winner)
}

##################
# sample_function
# input: election data, sample_size, num_candsn, tcheck
# output:  tibble(unique_winners_1, unique_winners_2, unique_winners_3, unique_winners_4,
#   unique_winners_max, unique_winners_noT1_1, unique_winners_noT1_2,
#   unique_winners_noT1_3, unique_winners_noT1_4, unique_winners_noT1_max)
#################

sample_function <- function(my_df, sample_size, num_candsn, tcheck) {
  num_unique_winners <- vector("integer", 1000)
  num_unique_winners_noT1 <- vector("integer", 1000)
  for (j in 1:1000) 
  {
    sample_data <- sample(my_df[[1]], sample_size, replace = TRUE, prob = my_df[[2]])
    sample_data_table <- tibble(sample_data) %>%
      transmute(ballot = sample_data,
                Count = 1)
#    Winner = IRV_function2(my_df = sample_data_table, num_cands = num_candsn)
    All_winners <- sapply(tcheck, IRV_Trunc_function2, sample_data_table, num_candsn)
    All_winners_noT1 <- All_winners[-1]
    num_unique_winners[[j]] <- length(unique(All_winners))
    num_unique_winners_noT1[[j]] <- length(unique(All_winners_noT1))
  }
  unique_winners_1 <- length(which(1 == num_unique_winners))
  unique_winners_2 <- length(which(2 == num_unique_winners))
  unique_winners_3 <- length(which(3 == num_unique_winners))
  unique_winners_4 <- length(which(4 == num_unique_winners))
  unique_winners_max <- max(num_unique_winners)
  unique_winners_noT1_1 <- length(which(1 == num_unique_winners_noT1))
  unique_winners_noT1_2 <- length(which(2 == num_unique_winners_noT1))
  unique_winners_noT1_3 <- length(which(3 == num_unique_winners_noT1))
  unique_winners_noT1_4 <- length(which(4 == num_unique_winners_noT1))
  unique_winners_noT1_max <- max(num_unique_winners_noT1)
  outputAll <- tibble(unique_winners_1, unique_winners_2, unique_winners_3, unique_winners_4,
           unique_winners_max, unique_winners_noT1_1, unique_winners_noT1_2,
           unique_winners_noT1_3, unique_winners_noT1_4, unique_winners_noT1_max)
  return(outputAll)
}

##################
# full_sample_function
# input: num_file
# output:  tibble(filename, unique_winners_1, unique_winners_2, unique_winners_3, unique_winners_4,
#   unique_winners_max, unique_winners_noT1_1, unique_winners_noT1_2,
#   unique_winners_noT1_3, unique_winners_noT1_4, unique_winners_noT1_max))
#################

full_sample_function <- function(num_file)
{
  data <- files[[num_file]]
  print(num_file)
  fixed_data <- Fix_data_function(data)
  my_df <- fixed_data[[1]]
  num_voters <- sum(my_df$Count)
  sample_size = min(1001, num_voters)
  num_cands <- fixed_data[[2]]
  truncs_to_check = 1:(num_cands-1)
  samplesAllData <- sample_function(my_df, sample_size, num_cands, truncs_to_check) %>%
    mutate(file = data) %>%
    select(file, everything())
  return(samplesAllData)
}

#######
# Make table of samples
######

ToOneFifty <- map_dfr(121:150, full_sample_function)

```

